---
title: "Linear Model Selection"
author: "Ruoqing Zhu"
date: "Last Updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    toc: yes
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: 2
---

<style>
body {
text-align: justify}
</style>

```{r set-options, echo=FALSE, cache=FALSE}
  options(width = 1000)
  knitr::opts_chunk$set(fig.width=9, fig.height=7, out.width = "75%", fig.align = 'center')
  knitr::opts_chunk$set(class.source = "fold-show")
  knitr::opts_chunk$set(collapse=TRUE)
```


## Model Selection Criteria

We will use the `diabetes` dataset from the `lars` package as a demonstration of model selection. Ten baseline variables include age, sex, body mass index, average blood pressure, and six blood serum measurements. These measurements were obtained for each of $n = 442$ diabetes patients, as well as the outcome of interest, a quantitative measure of disease progression one year after baseline. More details are available in the [`lars` package documentation](https://cran.r-project.org/web/packages/lars/lars.pdf). Our goal is to select a linear model, preferably with a small number of variables, that can predict the outcome. To select the best model, commonly used strategies include Marrow's $C_p$, AIC (Akaike information criterion) and BIC (Bayesian information criterion). 

```{r}
    # load the diabetes data
    library(lars)
    data(diabetes)
    diab = data.frame(cbind(diabetes$x, "Y" = diabetes$y))

    # fit linear regression with all covariates
    lm.fit = lm(Y~., data=diab)
```

The idea of model selection is to apply some penalty on the number of parameters used in the model. On one hand, we want to model fitting to be as good as possible, i.e., the mean squared error is small. On one hand, keeping adding covariates would always (theoretically) improve the model fitting (training error) and eventually give us a perfect fit. However, is this necessarily a good thing? We will end up with too many parameters, which is hard to interpret. But also, what about the prediction error? Are they necessarily better than a simple model? Let's investigate this with a simulation study.

## A Simulation Study

The idea of simulation is to assess the performance of a model by repeatedly generating data from a known underlying model. Usually, this requires a large number of repeats so that a pattern can be revealed. In the following example, we will see how the increasing number of covariates could affect the model performance. 

In __each simulation__, we will generate data and fit a linear regression. You should notice that only the first variable is useful. However, in practice, we may not know which one is truly useful. Hence, all variables will be used. 

```{r}
    set.seed(1)
    n = 100
    p = 20
    
    # the design matrix
    x = matrix(rnorm(n*p), n, p)
    
    # training data outcome
    ytrain = 0.3*x[, 1] + rnorm(n)
    
    # testing data outcome, using the same covariates
    ytest = 0.3*x[, 1] + rnorm(n)
    
    # construct the training and testing data
    traindata = data.frame("x" = x, "y" = ytrain)
    testdata = data.frame("x" = x, "y" = ytest)
    
    # fit model and calculate testing error
    onefit = lm(y~., data = traindata)
    ypred = predict(onefit, testdata)
    error = mean( (ypred - testdata$y)^2 )
    error
```

The question now is, if we use a smaller set of covariates, would the performance be better? We are going to do two things. First, we will try different model size, from 1 to 20. Second, we will repeat this process 100 times for each model size and take the averaged result. This can reasonably demonstrate the effect of using covariates. 

```{r}
    nsim = 100
    testerrors = matrix(NA, nsim, p)
    trainerrors = matrix(NA, nsim, p)
    
    for (i in 1:nsim)
    {
        # the design matrix
        x = matrix(rnorm(n*p), n, p)
        ytrain = 0.3*x[, 1] + rnorm(n)
        ytest = 0.3*x[, 1] + rnorm(n)

        # try all different dimensions
        for (j in 1:p)
        {
            # construct the data
            traindata = data.frame("x" = x[, 1:j, drop = FALSE], "y" = ytrain)
            testdata = data.frame("x" = x[, 1:j, drop = FALSE], "y" = ytest)
            # fit model and calculate testing error
            onefit = lm(y~., data = traindata)
            ypred = predict(onefit, testdata)
            error = mean( (ypred - testdata$y)^2 )

            testerrors[i, j] = error
            trainerrors[i, j] = mean( (onefit$fitted - traindata$y)^2 )
        }
    }

    plot(colMeans(testerrors), type = "b", 
         col = "darkorange", lwd = 3, 
         xlab = "Number of Variables", 
         ylab = "Prediction Error")
```

Hence, it seems that when only one variable is useful, adding any other variables would only increase the prediction error. The more irrelevant variables we add, the worse the result becomes. This is  over-fitting. On the other hand, the training error is still decreasing.

```{r}
    plot(colMeans(trainerrors), type = "b", 
         col = "darkorange", lwd = 3, 
         xlab = "Number of Variables", 
         ylab = "Training Error")
```

In many situations, variables are not completely useless. For example, in the following model, the effect of each variable decreases gradually:

\begin{align}
Y =& \mu + \epsilon \\
=& X^\text{T} \boldsymbol \beta + \epsilon \\
=& \sum_{j = 1}^p X_j 0.4^\sqrt{j} + \epsilon
\end{align}


```{r out.width='60%'}
    testerrors = matrix(NA, nsim, p)
    trainerrors = matrix(NA, nsim, p)

    for (i in 1:nsim)
    {
        # the design matrix
        x = matrix(rnorm(n*p), n, p)
        ytrain = x %*% 0.4^sqrt(c(1:p)) + rnorm(n)
        ytest = x %*% 0.4^sqrt(c(1:p)) + rnorm(n)
        
        # try all different dimensions
        for (j in 1:p)
        {
            # construct the data
            traindata = data.frame("x" = x[, 1:j, drop = FALSE], "y" = ytrain)
            testdata = data.frame("x" = x[, 1:j, drop = FALSE], "y" = ytest)
            
            # fit model and calculate testing error
            onefit = lm(y~., data = traindata)
            ypred = predict(onefit, testdata)
            error = mean( (ypred - testdata$y)^2 )
            
            testerrors[i, j] = error
            trainerrors[i, j] = mean( (onefit$fitted - traindata$y)^2 )
        }
    }
    
    plot(colMeans(testerrors), type = "b", 
         col = "darkorange", lwd = 3, 
         xlab = "Number of Variables", 
         ylab = "Prediction Error")
```

The training error will still decrease as expected. But we can see that the decreasing trend is not linear, meaning that for the first several variables, they have a stronger effect in explaining the outcome, and later variables are close to useless.

```{r}
    plot(colMeans(trainerrors), type = "b", 
         col = "darkorange", lwd = 3, 
         xlab = "Number of Variables", 
         ylab = "Training Error")
```

Combining these understandings, the best model seems to be using six variables, from the testing errors. The downside here is that we are not able to explain the signals of the remaining variables. However, the upside is that (from our first example), by using less number of variables (regardless of usefulness), we can improve the prediction error. This is essentially a __bias-variance trade-off__. We will introduce several methods that select variables and obtain a better model for prediction. Their essential ideas are similar, to introduce some penalty on the number of variables. 

$$\text{Goodness-of-Fit} + \text{Complexity Penality}$$

## Using Marrows' $C_p$

For example, the Marrows' $C_p$ criterion minimize the following quantity: 

$$\text{RSS} + 2 p \widehat\sigma_{\text{full}}^2$$
Note that the $\sigma_{\text{full}}^2$ refers to the residual variance estimation based on the full model, i.e., will all variables. Hence, this formula cannot be used when $p > n$ because you would not be able to obtain a valid estimation of $\sigma_{\text{full}}^2$. Nonetheless, we can calculate this quantity with the diabetes dataset

```{r}
    # number of variables (including intercept)
    p = 11
    n = nrow(diab)

    # obtain residual sum of squares
    RSS = sum(residuals(lm.fit)^2)

    # use the formula directly to calculate the Cp criterion 
    Cp = RSS + 2*p*summary(lm.fit)$sigma^2
    Cp
```

We can compare this with another sub-model, say, with just `age` and `glu`:

```{r}
    lm.fit_sub = lm(Y~ age + glu, data=diab)
  
    # obtain residual sum of squares
    RSS_sub = sum(residuals(lm.fit_sub)^2)
    
    # use the formula directly to calculate the Cp criterion 
    Cp_sub = RSS_sub + 2*3*summary(lm.fit)$sigma^2
    Cp_sub
```

Comparing this with the previous one, the full model is better. 

## Using AIC and BIC

Calculating the AIC and BIC criteria in `R` is a lot simpler, with the existing functions. The AIC score is given by 

$$-2 \text{Log-likelihood} + 2 p,$$
while the BIC score is given by 

$$-2 \text{Log-likelihood} + \log(n) p,$$

Interestingly, when assuming that the error distribution is Gaussian, the log-likelihood part is just a function of the RSS. In general, AIC performs similarly to $C_p$, while BIC tend to select a much smaller set due to the larger penalty. Theoretically, both AIC and $C_p$ are interested in the prediction error, regardless of whether the model is specified correctly, while BIC is interested in selecting the true set of variables, while assuming that the true model is being considered. 

The AIC score can be done using the `AIC()` function. We can match this result by writing out the normal density function and plug in the estimated parameters. Note that this requires one additional parameter, which is the variance. Hence the total number of parameters is 12. We can calculate this with our own code:

```{r}
    # ?AIC
    # a build-in function for calculating AIC using -2log likelihood
    AIC(lm.fit) 

    # Match the result
    n*log(RSS/n) + n + n*log(2*pi) + 2 + 2*p
```

Alternatively, the `extractAIC()` function can calculate both AIC and BIC. However, note that the `n + n*log(2*pi) + 2` part in the above code does not change regardless of how many parameters we use. Hence, this quantify does not affect the comparison between different models. Then we can safely remove this part and only focus on the essential ones. 

```{r}
    # ?extractAIC
    # AIC for the full model
    extractAIC(lm.fit)
    n*log(RSS/n) + 2*p

    # BIC for the full model
    extractAIC(lm.fit, k = log(n))
    n*log(RSS/n) + log(n)*p
```

Now, we can compare AIC or BIC using of two different models and select whichever one that gives a smaller value. For example the AIC of the previous sub-model is 

```{r}
    # AIC for the sub-model
    extractAIC(lm.fit_sub)
```

## Model Selection Algorithms

In previous examples, we have to manually fit two models and calculate their respective selection criteria and compare them. This is a rather tedious process if we have many variables and a huge number of combinations to consider. To automatically compare different models and select the best one, there are two common computational approaches: best subset regression and step-wise regression. As their name suggest, the best subset selection will exhaust all possible combination of variables, while the step-wise regression would adjust the model by adding or subtracting one variable at a time to reach the best model. 

## Best Subset Selection with `leaps`

Since the penalty is only affected by the number of variables, we may first choose the best model with the smallest RSS for each model size, and then compare across these models by attaching the penalty terms of their corresponding sizes. The `leaps` package can be used to calculate the best model of each model size. It essentially performs an exhaustive search, however, still utilizing some tricks to skip some really bad models. Note that the `leaps` package uses the data matrix directly, instead of specifying a formula.  

```{r}
    library(leaps)
    
    # The package specifies the X matrix and outcome y vector
    RSSleaps = regsubsets(x = as.matrix(diab[, -11]), y = diab[, 11])
    summary(RSSleaps, matrix=T)
```

The results is summarized in a matrix, with each row representing a model size. The `"*"` sign indicates that the variable is include in the model for the corresponding size. Hence, there should be only one of such in the first row, two in the second row, etc. 
    
By default, the algorithm would only consider models up to size 8. This is controlled by the argument `nvmax`. If we want to consider larger model sizes, then set this to a larger number. However, be careful that this many drastically increase the computational cost. 

```{r}    
    # Consider maximum of 10 variables
    RSSleaps = regsubsets(x = as.matrix(diab[, -11]), y = diab[, 11], nvmax = 10)
    summary(RSSleaps,matrix=T)
    
    # Obtain the matrix that indicates the variables
    sumleaps = summary(RSSleaps, matrix = T)
    
    # This object includes the RSS results, which is needed to calculate the scores
    sumleaps$rss
    
    # This matrix indicates whether a variable is in the best model(s)
    sumleaps$which
    
    # The package automatically produces the Cp statistic
    sumleaps$cp
```

We can calculate different model selection criteria with the best models of each size. The model fitting result already produces the $C_p$ and BIC results. However, please note that both quantities are modified slightly. For the $C_p$ statistics, the quantity is divided by the estimated error variance, and also adjust for the sample size. For the BIC, the difference is a constant regardless of the model size. Hence these difference do will not affect the model selection result because the modification is the same regardless of the number of variables. 

```{r}
    modelsize=apply(sumleaps$which,1,sum)
    
    Cp = sumleaps$rss/(summary(lm.fit)$sigma^2) + 2*modelsize - n;
    AIC = n*log(sumleaps$rss/n) + 2*modelsize;
    BIC = n*log(sumleaps$rss/n) + modelsize*log(n);
    
    # Comparing the Cp scores 
    cbind("Our Cp" = Cp, "leaps Cp" = sumleaps$cp) 
    
    # Comparing the BIC results. The difference is a constant, 
    # which is the score of an intercept model
    cbind("Our BIC" = BIC, "leaps BIC" = sumleaps$bic, 
          "Difference" = BIC-sumleaps$bic, 
          "Intercept Score" = n*log(sum((diab[,11] - mean(diab[,11]))^2/n)))
```

Finally, we may select the best model, using any of the criteria. The following code would produced a plot to visualize it. We can see that BIC selects 6 variables, while both AIC and $C_p$ selects 7. 

```{r}
    # Rescale Cp, AIC and BIC to (0,1).
    inrange <- function(x) { (x - min(x)) / (max(x) - min(x)) }
    
    Cp = inrange(Cp)
    BIC = inrange(BIC)
    AIC = inrange(AIC)

    plot(range(modelsize), c(0, 0.4), type="n", 
         xlab="Model Size (with Intercept)", 
         ylab="Model Selection Criteria", cex.lab = 1.5)

    points(modelsize, Cp, col = "green4", type = "b", pch = 19)
    points(modelsize, AIC, col = "orange", type = "b", pch = 19)
    points(modelsize, BIC, col = "purple", type = "b", pch = 19)
    legend("topright", legend=c("Cp", "AIC", "BIC"),
           col=c("green4", "orange", "purple"), 
           lty = rep(1, 3), pch = 19, cex = 1.7)
```

## Step-wise regression using `step()`

The idea of step-wise regression is very simple: we start with a certain model (e.g. the intercept or the full mode), and add or subtract one variable at a time by making the best decision to improve the model selection score. The `step()` function implements this procedure. The following example starts with the full model and uses AIC as the selection criteria (default of the function). After removing several variables, the model ends up with six predictors.  

```{r}
    # k = 2 (AIC) is default; 
    step(lm.fit, direction="both", k = 2)
```

We can also use different settings, such as which model to start with, which is the minimum/maximum model, and do we allow to adding/subtracting. 

```{r}
    # use BIC (k = log(n))instead of AIC
    # trace = 0 will suppress the output of intermediate steps 
    step(lm.fit, direction="both", k = log(n), trace=0)

    # Start with an intercept model, and use forward selection (adding only)
    step(lm(Y~1, data=diab), scope=list(upper=lm.fit, lower=~1), 
         direction="forward", trace=0)
```

We can see that these results are slightly different from the best subset selection. So which is better? Of course the best subset selection is better because it considers all possible candidates, which step-wise regression may stuck at a sub-optimal model, while adding and subtracting any variable do not benefit further. Hence, the results of step-wise regression may be unstable. On the other hand, best subset selection not really feasible for high-dimensional problems because of the computational cost. 