---
title: "Stat 432 Homework 1"
date: "Assigned: Aug 26, 2024; <span style='color:red'>Due: 11:59 PM CT, Sep 5, 2024</span>"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: 2
---

<style>
  body {
  text-align: justify}
</style>

```{css, echo=FALSE}
  .solution {
  background-color: #CCDDFF; /* Replace with your desired color */
  }
  
  blockquote {
  background-color: #CCDDFF; /* Replace with your desired color */
  font-family: "Times New Roman", serif; /* Change to your desired font family */
  /* font-size: 16px; /* Change to your desired font size */
  font-weight: bold; /* Makes the text bold */
  }
```

## Instruction

<span style='color:red'>__Please remove this section when submitting your homework.__</span>

Students are encouraged to work together on homework and/or utilize advanced AI tools. However, __sharing, copying, or providing any part of a homework solution or code to others__ is an infraction of the [University's rules on Academic Integrity](https://studentcode.illinois.edu/article1/part4/1-401/). Any violation will be punished as severely as possible. Final submissions must be uploaded to [Gradescope](https://www.gradescope.com/courses/570816). No email or hard copy will be accepted. For [**late submission policy and grading rubrics**](https://teazrq.github.io/stat432/syllabus.html), please refer to the course website.

- You are required to submit the rendered file `HWx_yourNetID.pdf`. For example, `HW01_rqzhu.pdf`. Please note that this must be a `.pdf` file. `.html` format __cannot__ be accepted. Make all of your `R` code chunks visible for grading. 
- Include your Name and NetID in the report.
- If you use this file or the example homework `.Rmd` file as a template, be sure to __remove this instruction__ section.
- Make sure that you __set seed__ properly so that the results can be replicated if needed.
- For some questions, there will be restrictions on what packages/functions you can use. Please read the requirements carefully. As long as the question does not specify such restrictions, you can use anything.
- __When using AI tools__, try to document your prompt and any follow-up prompts that further modify or correct the answer. You are also required to briefly comment on your experience with it, especially when it's difficult for them to grasp the idea of the question.
- __On random seed and reproducibility__: Make sure the version of your `R` is $\geq 4.0.0$. This will ensure your random seed generation is the same as everyone else. Please note that updating the `R` version may require you to reinstall all of your packages. 

## Question 1 (Multivariate Normal Distribution)

This question is about playing with AI tools for generating multivariate normal random variables. Let $X_i$, $i = 1, \ldots, n$ be i.i.d. multivariate normal random variables with mean $\mu$ and covariance matrix $\Sigma$, where 
$$ 
\mu = \begin{bmatrix} 1 \\ 2 \end{bmatrix}, \quad \text{and} \quad \Sigma = \begin{bmatrix} 1 & 0.5 \\ 0.5 & 1 \end{bmatrix}.
$$
Write `R` code to perform the following tasks. Please try to use AI tools as much as possible in this question. 

a. [10 points] Generate a set of $n = 2000$ observations from this distribution. Only display the first 5 observations in your `R` output. Make sure set random seed $=1$ in order to replicate the result. Calculate the sample covariance matrix of the generated data and compare it with the true covariance matrix $\Sigma$.

b. [10 points] If you used VS Code and AI tools to perform the previous question, then they will most likely suggest using the `mvrnorm` function from the `MASS` package. However, there are alternative ways to complete this question. For example, you could first generate $n$ standard normal random variables, and then transform them to the desired distribution. Write down the mathematical formula of this approach in Latex, and then write `R` code to implement this approach. Only display the first 5 observations in your `R` output. Validate your approach by computing the sample covariance matrix of the generated data and compare it with the true covariance matrix $\Sigma$. Please note that you __should not use__ the `mvrnorm` function anymore in this question. 

c. [10 points] Write an `R` function called `mymvnorm` that takes the following arguments: `n`, `mu`, `sigma`. The function should return a matrix of dimension $n \times p$, where $p$ is the length of `mu`. The function should generate $n$ observations from a multivariate normal distribution with mean `mu` and covariance matrix `sigma`. You should not use the `mvrnorm` function in your code. Instead, use the logic you wrote in part b) to generate the data. Again, validate your result by calculating the sample covariance matrix of the generated data and compare to $\Sigma$. Also, when setting seed correctly, your answer in this question should be identical to the one in part b).

d. [10 points] If you used any AI tools during the first three questions, write your experience here. Specifically, what tool(s) did you use? __What prompt was used__? Did the tool suggested a corrected answer to your question? If not, which part was wrong? How did you corrected their mistakes (e.g modifying your prompt)?

## Question 2 (Data Manipulation Plots)

The following question practices data manipulation and summary statistics. Our goal is to write a function that calculates the price gap between any two given dates. Load the `quantmod` package and obtain the `AAPL` data (apple stock price). 

```{r class.source="solution", warning=FALSE, message=FALSE, fig.align='center', out.width='60%'}
  library(quantmod)
  getSymbols("AAPL")
  plot(AAPL$AAPL.Close, pch = 19)
```

a. [20 points] Calculate a 90-day moving average of the closing price of `AAPL` and plot it on the same graph. Moving average means that for each day, you take the average of the previous 90 days. Please do this in two ways: 1) there is a built-in function called `SMA` in the `quantmod` package; 2) write your own function to calculate the moving average. For both questions, you can utilize AI tools to help you write the code.

b. [15 points] I have an alternative way of writing this function. 

```{r class.source="solution", fig.align='center', out.width='60%'}
  my_average <- function(x, window) {
    # Compute the moving average of x with window size = window
    n <- length(x)
    ma <- rep(NA, n)
    for (i in window:n) {
      myinterval = (i-window/2):(i + window/2)
      myinterval = myinterval[myinterval > 0 & myinterval <= n]
      ma[i] <- mean( x[ myinterval ] )
    }
    return(ma)
  }

  AAPL$MA90 <- my_average(Cl(AAPL), 90)
  plot(AAPL$AAPL.Close, pch = 19)
  lines(AAPL$MA90, col = "red", lwd = 3)
```

Can you comment on the difference of these two functions? Do you think my line is a good choice when it is used for predicting future prices? Which one do you prefer and why.

## Question 3 (Read/write Data)

a. [10 Points] The `ElemStatLearn` package [[CRAN link](https://cran.r-project.org/web/packages/ElemStatLearn/index.html)] is an archived package. Hence, you cannot directly install it using the `install.packages()` function. Instead, you may install an older version of it by using the `install_github()` function from the `devtools` package. Install the `devtools` package and run the find the code to install the `ElemStatLearn` package.

b. [15 Points] Load the `ElemStatLearn` package and obtain the `ozone` data. Save this data into a `.csv` file, and then read the data back from that file into `R`. Print out the first 5 observations to make sure that the new data is the same as the original one.